Document Indexing Using Aho-Corasick State Machines

### 2011-10-20 report/article.tex
Paper down to 11 pages - sending for review

\documentclass[10pt]{article}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}

\title{Document Indexing Aho-Corasick State Machines}
\author{Damien DuBois, Justin Kamerman, Ramanpreet Singh}
\date{\today}

@@

%----------------------------------------
% Abstract
%----------------------------------------
\begin{abstract}
The Aho-Corasick algorithm was originally proposed as a bibliographic
search mechanism, efficient enough to preclude the construction and
maintenance of a search index. As the size of document collections
have grown since the development of the algorithm, document
collections have grown to the point where cost of rescanning a corpus
for every search has become prohibitive and indexing, in some form or
another, is an essential optimization for any modern information
retrieval system. In this paper we demonstrate the use of
Aho-Corasick, not as an alternative to indexing but as a lexical
scanning tool employed in the construction of an inverted index. We
provide analytic predictions of our indexing algorithm and an
empirical evaluation of our inverted index implementation is conducted
and the results analyzed. Consistent with our analysis, the hybrid
approach yields significant improvement over a naive
implementation. In addition, the modular architecture of our indexer
allowed the indexing task to be parallelized over multiple physical
nodes. The results of parallelization are presented and offer
interesting insights towards further horizontal scaling.
\\\\
{\bf Keywords:} inverted index, Aho-Corasick state machine, finite
automata, pattern matching, parsing
\end{abstract}

%----------------------------------------
% Introduction
%----------------------------------------
\section{Introduction}
Information Retrieval is a fast becoming the dominant form of
information access, overtaking traditional relational database style
searching\cite{RefWorks:109}. The basis of most document retrieval systems is the
term-document incidence matrix. This matrix is typically sparse and
more efficiently represented as an inverted-index which maps terms to
the parts of a document where they occur.

@@

%----------------------------------------
% Research Plan
%----------------------------------------
\section{Research Plan}
\label{sec:researchplan}
The primary goals of this research project is to construct an
information retrieval framework within which specific implementations
can be evaluated under various operating conditions. To initiate the
synthesis of this general framework, a reference implementation will
be built, based on the notion of a simple inverted index.

The software architecture of the reference implementation will be
carefully documented and follow established design patterns in order
to facilitate the future construction of alternative document search
schemes within the same framework. The framework will be designed to
be executed in a distributed deployment such that it may be used to
evaluate the effect of parallelization on a particular algorithm.

The performance of the reference implementation will be tested with
respect to the following operating parameters and compared to analytic
expectations:

\begin{itemize}
  \item Number of index keywords
  \item Number of documents in collection
  \item Average size of documents
\end{itemize}

In addition, the reference implementation will be compared against a
naïve implementation of the inverted indexer. The naïve implementation
provided a baseline for further evaluation.

%----------------------------------------
% Design
%----------------------------------------
\section{Design}
\label{sec:design}
A black-box functional view of the indexing system is show in figure
\ref{fig:blackbox}. The system processes a continuous incoming flow of
documents and distribute them to parallel indexing threads running on
physically separate, heterogeneous nodes.  Document queries are performed
using an evolving search index. The operational balance is to
timeously index new additions to the corpus while servicing concurrent
search requests. A view as to how the components of the indexer system
are deployed is show in figure \ref{fig:deploymentmodel}.

As can be seen in figure \ref{fig:deploymentmodel}, the indexer
processes are symmetrical and deployed on multiple physical nodes. The
individual indexers do not interact directly with one another, making
for a simple deployment and operation model. The execution loop of
each indexer is as follows:

@@

\begin{figure}
  \begin{center}
        \includegraphics[width=\textwidth,height=!]{blackbox}
  \end{center}
  \caption{A black-box view of the indexer}
  \label{fig:blackbox}
\end{figure}

\begin{figure}
  \begin{center}
        \includegraphics[width=\textwidth,height=!]{deploymentmodel}
  \end{center}
  \caption{Indexer deployment model}
  \label{fig:deploymentmodel}
\end{figure}

@@

\subsubsection{Time Complexity}
\begin{itemize}
\item \textbf{State Machine Construction:}
  The time taken to construct the Aho-Corasick state machine was
  measured for different numbers of keywords. The results of this test
  is shown in figure \ref{fig:numkeywordstimecomplexbuildsm} and construction time
  can easily be considered linear with respect to the number of
  keywords. This is consistent with \cite{RefWorks:103} which proves
  that the state machine construction algorithm is linearly
  proportional to the sum of the lengths of the keywords used to
  construct the state machine.

  \begin{figure}[h]
    \begin{center}
      \includegraphics[width=\textwidth,height=!]{numkeywordstimecomplexbuildsm}
    \end{center}
    \caption{Time to build Aho-Corasick state machine as a function of
      the number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
    \label{fig:numkeywordstimecomplexbuildsm}
  \end{figure}

\item \textbf{Index Construction:} 
  The time taken for a state machines to build an inverted index of our
  corpus was measured. The test was repeated using state machines
  constructed with a varying number of keywords and the results plotted
  in figure \ref{fig:numkeywordstimecomplexbuildindex}. According to
  \cite{RefWorks:103} which proves that the number of state transitions
  involved in processing an input string is independent of the number of
  keywords used to construct the state machine, we would not expect the
  time complexity to increase with the size of the keyword set. However,
  as can be seen in \ref{fig:numkeywordstimecomplexbuildindex}, the time
  to build the index increases logarithmically with the number of
  keywords. This is likely due to the fact that as the number of keywords
  increases, so does the size of the index and the number of database
  interacts required to store and update the index.

  \begin{figure}[h]
    \begin{center}
      \includegraphics[width=\textwidth,height=!]{numkeywordstimecomplexbuildindex}
    \end{center}
    \caption{Time to build the inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
    \label{fig:numkeywordstimecomplexbuildindex}
  \end{figure}

\end{itemize}

\subsubsection{Space Complexity}
As we would have expected it, the space used by the index in the
database is proportional to the number of words we are using. It also
means that our words have the same probability at the beginning or at
the end of our 12,000 word dictionary. This result is reflected in
figure \ref{fig:numkeywordsspacecomplexbuildindex}.

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{numkeywordsspacecomplexbuildindex}
  \end{center}
    \caption{Size of index as a function of the number of
      keywords. The number of database rows used to store the index is
      shown on the y-axis and the number of keywords on the x-axis} 
    \label{fig:numkeywordsspacecomplexbuildindex}
\end{figure}

@@

\subsubsection{Time Complexity}
As we were expecting, the time to build the index with the state
machine strategy is linear with respect to the number of post, as
depicted in figure \ref{fig:corpsizetimecomplexbuildindex}. However, to be able
to perform this experiment we had to run the indexer program with
batches of 10,000 posts due to Java heap space problems. As a result, the state
machine is rebuilt every 10,000 posts.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{corpsizetimecomplexbuildindex}
  \end{center}
    \caption{Time to build the inverted index as a function of the
      size of the corpus. Time taken is shown in milliseconds on the
      y-axis and the number of documents in the collection on the
      x-axis} 
    \label{fig:corpsizetimecomplexbuildindex}
\end{figure}

\subsubsection{Space Complexity}
As expected, the space used to store the index in the database is
proportional to the size of the corpus. This result is shown in figure
\ref{fig:corpsizespacecomplexbuildindex}.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{corpsizespacecomplexbuildindex}
  \end{center}
    \caption{Size of index as a function of the size of the
      corpus. The number of database rows used to store the index is
      shown on the y-axis and the number of documents in the
      collection on the x-axis} 
    \label{fig:corpsizespacecomplexbuildindex}
\end{figure}

@@

\subsubsection{Time Complexity}
\begin{itemize}
\item \textbf{State Machine Construction:} 
  As expected, state machine construction time
  does not depend on the length of the documents. Nonetheless it is
  interesting to compare figures \ref{fig:docsizetimecomplexbuildsm}
  and \ref{fig:docsizetimecomplexbuildindex}, showing the construction
  times for the state machine and inverted index relative to the
  average document size. The comparison shows that state machine
  construction time is several orders of magnitude smaller than index
  construction and negligible by comparison. Our state machine is
  built in 770 ms on average for 1000 keywords with a standard
  deviation of 18ms.

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=\textwidth,height=!]{docsizetimecomplexbuildsm}
    \end{center}
    \caption{Time to build Aho-Corasick state machine as a function of
      the average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis}
    \label{fig:docsizetimecomplexbuildsm}
  \end{figure}

\item \textbf{Index Construction:} 
  As expected the time to build the indexer is almost linear with
  respect to the average document size. This relationship is shown in
  figure \ref{fig:docsizetimecomplexbuildindex}.

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=\textwidth,height=!]{docsizetimecomplexbuildindex}
    \end{center}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis} 
    \label{fig:docsizetimecomplexbuildindex}
  \end{figure}

\end{itemize}

\subsubsection{Space Complexity}
Here we are studying the number of records in our index database as a
function of the average document size. As expected, the larger the documents
the more records created to represent the index. However, this is not linear
since if a word occurs many times in the same post we just save it
once in the database. This relationship is shown in figure
\ref{fig:docsizespacecomplexbuildindex}.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{docsizespacecomplexbuildindex}
  \end{center}
  \caption{Size of index as a function of the average document
    size. The number of database rows used to store the index is
      shown on the y-axis and the average document size in characters
      on the x-axis}  
  \label{fig:docsizespacecomplexbuildindex}
\end{figure}

@@

\subsubsection{Complexity Analysis}
For the na\"{i}ve algorithm described above, the theoretical worst
case time complexity relationship is shown in equation
\ref{eqn:timecomplexworstcase}.

@@

\subsubsection{Document Size}
For this experiment we fixed the number of keywords to 1,000 and the
number of documents to 5,000 and we ran our code on a database
composed of document of different length. The experiment results are
shown in figure \ref{fig:naivedocumentsize}.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivedocumentsize}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in miliseconds on the
      y-axis and the average document size in characters on the x-axis}
  \label{fig:naivedocumentsize}
\end{figure}

As expected, the curves for both our indexer and the naive indexer are
almost linear since for the naive implementation:

@@

We can observe that for a small length of document naive method is
faster since \(keyNum\) is smaller than \(docNum\). For \(docLength = 1300\)
our indexer is far more efficient. To see this effect more clearly,
experiment results are plotted on a logarithmic scale in figure
\ref{fig:naivedocumentsizelog}.

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivedocumentsizelog}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the
      logarithmic x-axis}
  \label{fig:naivedocumentsizelog}
\end{figure}

\subsubsection{Number of Keywords}
As expected, the two implementations exhibit a linear response to the
number of keywords used. In the na\"{i}ve implemementation:

@@

\begin{figure}[p]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivenumkeywords}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
  \label{fig:naivenumkeywords}
\end{figure}

\begin{figure}[p]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivenumkeywordslog}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the logarithmic x-axis}
  \label{fig:naivenumkeywordslog}
\end{figure}

\subsubsection{Size of Corpus}
As expected, the time complexity is linearly related to the number of
documents being indexed. The experiment results are shown in figure
\ref{fig:naivesizecorpus} and on a logarithmic scale in figure
\ref{fig:naivesizecorpuslog}. For the na\"{i}ve implementation:

@@

\begin{figure}[p]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivesizecorpus}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      number of documents being indexed. Time taken is shown in
      milliseconds on the y-axis and the number of documents on the
      x-axis} 
  \label{fig:naivesizecorpus}
\end{figure}

\begin{figure}[p]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{naivesizecorpuslog}
  \end{center}
  \caption{Time to build an inverted index as a function of the
      number of documents being indexed. Time taken is shown in
      milliseconds on the y-axis and the number of documents on the
      logarithmic x-axis} 
  \label{fig:naivesizecorpuslog}
\end{figure}

%----------------------------------------
% Parallelization
%----------------------------------------
\subsection{Parallelization}
\label{parallelization}
Given the modular architecture of our indexer, it is possible to
deploy multiple instances of the program to work concurrently on the same
indexing task. The instances interact via the database which
synchronizes their operations on the corpus and the index itself. To
investigate the effect of parallelization on the indexer operation, the
indexer was deployed as follows:

\begin{itemize}
\item Single database server storing documents, keywords, and the
  resulting index.
\item Four indexer nodes each running a single instance of the indexer
  program. 
\end{itemize}

This deployment as well as a description of the sequence of operation
are shown in figure \ref{fig:paralleldeploymentmodel}.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{paralleldeploymentmodel}
  \end{center}
  \caption{Deployment model of parallel indexer} 
  \label{fig:paralleldeploymentmodel}
\end{figure}

In this experiment a corpus of 100,000 documents and 1000 keywords were
used, each document 5000 words in length. Tests were run using various
node combinations and the time taken to index the corpus was measured and
normalized to a throughput metric (documents/second) for convenient
comparison. Test results using the same number of nodes were averaged
to account for differences in the physical capabilities and
configuration of the worker nodes. Next, a series of tests was
performed using a corpus of 800,000 documents to investigate the
effect of corpus size on performance. The results of this experiment
are shown in figure \ref{fig:parallelization} and described in detail
in the following section.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=\textwidth,height=!]{parallelization}
  \end{center}
  \caption{Performance characteristics of parallel indexer
    deployment. The \textit{ideal} curve shows the theoretical
    performance curve based on linear combination of average
    individual node performance over the corpus} 
  \label{fig:parallelization}
\end{figure}

\subsubsection{Results}
For the 100,000 document corpus, increasing the number of nodes
increases throughput but with diminishing returns. As the database
connection pool is saturated, this becomes the limiting factor of the
architecture. If the number of nodes were increased further, one would
expect the curve to plateau at some point. After this point, adding
additional nodes would offer no improvement and may even decrease
performance as the worker processes spend an increasing amount of time
context switching and contending for locks. These factors could be
mitigated up to a point by increasing batch sizes and tuning the
database. However, the characteristics of the curve are ultimately
unavoidable in the current architecture.

@@

>>>>

\documentclass[10pt]{article}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{wrapfig}
\usepackage{caption3} % load caption package kernel first
\DeclareCaptionOption{parskip}[]{} % disable "parskip" caption option
\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}

\title{Document Indexing Using Aho-Corasick State Machines}
\author{Damien DuBois, Justin Kamerman, Ramanpreet Singh}
\date{\today}

@@

%----------------------------------------
% Abstract
%----------------------------------------
\begin{abstract}
The Aho-Corasick algorithm was originally proposed as a bibliographic
search mechanism, efficient enough to preclude the construction and
maintenance of a search index. As the size of document collections
have grown since the development of the algorithm, document
collections have grown to the point where cost of rescanning a corpus
for every search has become prohibitive and indexing, in some form or
another, is an essential optimization for any modern information
retrieval system. In this paper we demonstrate the use of
Aho-Corasick, not as an alternative to indexing but as a lexical
scanning tool employed in the construction of an inverted
index. Consistent with analytical predictions, an empirical evaluation
of our hybrid indexer, shows significant improvement over a
na\"{\i} implementation. In addition, the modular architecture of our indexer
allowed the indexing task to be parallelized over multiple physical
nodes. The results of parallelization are presented and offer
interesting insights towards further horizontal scaling.
\\\\
{\bf Keywords:} inverted index, Aho-Corasick state machine, finite
automata, pattern matching, parsing
\end{abstract}

%----------------------------------------
% Introduction
%----------------------------------------
\section{Introduction}
The basis of most document retrieval systems is the
term-document incidence matrix. This matrix is typically sparse and
more efficiently represented as an inverted-index which maps terms to
the parts of a document where they occur.

@@

%----------------------------------------
% Design
%----------------------------------------
\section{Implementation}
\label{sec:implementation}
As can be seen in figure \ref{fig:deploymentmodel}, the indexer
processes are symmetrical and can deployed on multiple physical nodes. The
individual indexers do not interact directly with one another, making
for a simple deployment and operation model. The execution loop of
each indexer is as follows:

@@

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{deploymentmodel}
  \end{center}
  \caption{Indexer deployment model}
  \label{fig:deploymentmodel}
\end{wrapfigure}

@@

The time taken to construct the Aho-Corasick state machine was
measured for different numbers of keywords. The results of this test
is shown in figure \ref{fig:numkeywordstimecomplexbuildsm} and
construction time can easily be considered linear with respect to the
number of keywords. This is consistent with \cite{RefWorks:103} which
proves that the state machine construction algorithm is linearly
proportional to the sum of the lengths of the keywords used to
construct the state machine.

The time taken for a state machines to build an inverted index of our
corpus was measured. The test was repeated using state machines
constructed with a varying number of keywords and the results plotted
in figure \ref{fig:numkeywordstimecomplexbuildindex}. According to
\cite{RefWorks:103} which proves that the number of state transitions
involved in processing an input string is independent of the number of
keywords used to construct the state machine, we would not expect the
time complexity to increase with the size of the keyword set. However,
as can be seen in \ref{fig:numkeywordstimecomplexbuildindex}, the time
to build the index increases logarithmically with the number of
keywords. This is likely due to the fact that as the number of
keywords increases, so does the size of the index and the number of
database interacts required to store and update the index.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{numkeywordstimecomplexbuildsm}
    \caption{Time to build Aho-Corasick state machine as a function of
      the number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
    \label{fig:numkeywordstimecomplexbuildsm}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{numkeywordstimecomplexbuildindex}
    \caption{Time to build the inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
    \label{fig:numkeywordstimecomplexbuildindex}
  \end{minipage}
\end{figure}

The space used by the index in the database is proportional to the
number of words we are using. It also means that our words have the
same probability at the beginning or at the end of our 12,000 word
dictionary.

@@

As we were expecting, the time to build the index with the state
machine strategy is linear with respect to the number of post, as
depicted in figure \ref{fig:corpsizetimecomplexbuildindex}. However, to be able
to perform this experiment we had to run the indexer program with
batches of 10,000 posts due to Java heap space problems. As a result, the state
machine is rebuilt every 10,000 posts.

The space used to store the index in the database is
proportional to the size of the corpus as shown in figure
\ref{fig:corpsizespacecomplexbuildindex}.

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{corpsizetimecomplexbuildindex}
    \caption{Time to build the inverted index as a function of the
      size of the corpus. Time taken is shown in milliseconds on the
      y-axis and the number of documents in the collection on the
      x-axis} 
    \label{fig:corpsizetimecomplexbuildindex}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{corpsizespacecomplexbuildindex}
    \caption{Size of index as a function of the size of the
      corpus. The number of database rows used to store the index is
      shown on the y-axis and the number of documents in the
      collection on the x-axis} 
    \label{fig:corpsizespacecomplexbuildindex}
  \end{minipage}
\end{figure}

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizetimecomplexbuildsm}
    \caption{Time to build Aho-Corasick state machine as a function of
      the average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis}
    \label{fig:docsizetimecomplexbuildsm}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{docsizetimecomplexbuildindex}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the x-axis} 
    \label{fig:docsizetimecomplexbuildindex}
  \end{minipage}
\end{figure}

State machine construction time does not depend on the
length of the documents. Nonetheless it is interesting to compare
figures \ref{fig:docsizetimecomplexbuildsm} and
\ref{fig:docsizetimecomplexbuildindex}, showing the construction times
for the state machine and inverted index relative to the average
document size. The comparison shows that state machine construction
time is several orders of magnitude smaller than index construction
and negligible by comparison. Our state machine is built in 770 ms on
average for 1000 keywords with a standard deviation of 18ms.

As expected the time to build the indexer is almost linear with
respect to the average document size. This relationship is shown in
figure \ref{fig:docsizetimecomplexbuildindex}.

As expected, the larger the documents the more records created to
represent the index. However, the relationship is not linear since if
a word occurs many times in the same post we just save it once in the
database. This relationship is shown in figure
\ref{fig:docsizespacecomplexbuildindex}.

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.70\textwidth,height=!]{docsizespacecomplexbuildindex}
  \end{center}
  \caption{Size of index as a function of the average document
    size. The number of database rows used to store the index is
    shown on the y-axis and the average document size in characters
    on the x-axis}  
  \label{fig:docsizespacecomplexbuildindex}
\end{wrapfigure}

@@

For the na\"{i}ve algorithm described above, the theoretical worst
case time complexity relationship is shown in equation
\ref{eqn:timecomplexworstcase}.

@@

To exmaine the effect of document size, the number of keywords was
fixed at 1,000 and the number of documents to 5,000 and we ran our
code on a database composed of document of different length. The
experiment results are shown in figure \ref{fig:naivedocumentsize}. As
expected, the curves for both our indexer and the naive indexer are
almost linear since for the naive implementation:

@@

We can observe that for a small length of document naive method is
faster since \(keyNum\) is smaller than \(docNum\). For \(docLength = 1300\)
our indexer is far more efficient. To see this effect more clearly,
experiment results are plotted on a logarithmic scale in figure
\ref{fig:naivedocumentsizelog}.As expected, the two implementations
exhibit a linear response to the number of keywords used. In the
na\"{i}ve implemementation:

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivedocumentsize}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in miliseconds on the
      y-axis and the average document size in characters on the x-axis}
    \label{fig:naivedocumentsize}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivedocumentsizelog}
    \caption{Time to build an inverted index as a function of the
      average document size. Time taken is shown in milliseconds on the
      y-axis and the average document size in characters on the
      logarithmic x-axis}
    \label{fig:naivedocumentsizelog}
  \end{minipage}
\end{figure}

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivenumkeywords}
    \caption{Time to build an inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the x-axis}
    \label{fig:naivenumkeywords}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivenumkeywordslog}
    \caption{Time to build an inverted index as a function of the
      number of keywords. Time taken is shown in milliseconds on the
      y-axis and the number of keywords on the logarithmic x-axis}
    \label{fig:naivenumkeywordslog}
  \end{minipage}
\end{figure}

As expected, the time complexity is linearly related to the number of
documents being indexed. The experiment results, in this respect, are
shown in figure \ref{fig:naivesizecorpus} and on a logarithmic scale
in figure \ref{fig:naivesizecorpuslog}. For the na\"{i}ve
implementation:

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivesizecorpus}
    \caption{Time to build an inverted index as a function of the
      number of documents being indexed. Time taken is shown in
      milliseconds on the y-axis and the number of documents on the
      x-axis} 
    \label{fig:naivesizecorpus}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=\textwidth]{naivesizecorpuslog}
    \caption{Time to build an inverted index as a function of the
    number of documents being indexed. Time taken is shown in
      milliseconds on the y-axis and the number of documents on the
      logarithmic x-axis} 
  \label{fig:naivesizecorpuslog}
  \end{minipage}
\end{figure}

%----------------------------------------
% Parallelization
%----------------------------------------
\subsection{Parallelization}
\label{parallelization}
Given the modular architecture of our indexer, it is possible to
deploy multiple instances of the program to work concurrently on the same
indexing task. The instances interact via the database which
synchronizes their operations on the corpus and the index itself. To
investigate the effect of parallelization on the indexer operation, the
indexer was deployed using a single database server storing documents,
keywords, and the resulting index; four indexer nodes each running a
single instance of the indexer program.

In this experiment a corpus of 100,000 documents and 1000 keywords were
used, each document 5000 words in length. Tests were run using various
node combinations and the time taken to index the corpus was measured and
normalized to a throughput metric (documents/second) for convenient
comparison. Test results using the same number of nodes were averaged
to account for differences in the physical capabilities and
configuration of the worker nodes. Next, a series of tests was
performed using a corpus of 800,000 documents to investigate the
effect of corpus size on performance. The results of this experiment
are shown in figure \ref{fig:parallelization}.

For the 100,000 document corpus, increasing the number of nodes
increases throughput but with diminishing returns. As the database
connection pool is saturated, this becomes the limiting factor of the
architecture. If the number of nodes were increased further, one would
expect the curve to plateau at some point. After this point, adding
additional nodes would offer no improvement and may even decrease
performance as the worker processes spend an increasing amount of time
context switching and contending for locks. These factors could be
mitigated up to a point by increasing batch sizes and tuning the
database. However, the characteristics of the curve are ultimately
unavoidable in the current architecture.

@@

\begin{wrapfigure}{r}{0.70\textwidth}
  \begin{center}
        \includegraphics[width=0.60\textwidth,height=!]{parallelization}
  \end{center}
  \caption{Performance characteristics of parallel indexer
    deployment. The \textit{ideal} curve shows the theoretical
    performance curve based on linear combination of average
    individual node performance over the corpus} 
\end{wrapfigure}
